---
title: "Covid-19"
author: "G. Ubit"
date: "2025-05-13"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

**About the data:**
The dataset is a global COVID19 dataset gotten from the Johns Hopkins github site found at <https://github.com/CSSEGISandData/COVID-19/tree/master/csse_covid_19_data/csse_covid_19_time_series>

**Statement of the question of interest:**
I am interested in understanding the relationship between Covid-19 cases per million and deaths per million globally and in Nigeria.

```{r everything}
#Libraries used in analysis
library(stringr)
library(tidyverse)
library(dplyr)
library(ggplot2)
library(readr)
library(utils)
library(patchwork)


## Input file from the 
url_in = "https://raw.githubusercontent.com/CSSEGISandData/COVID-19/master/csse_covid_19_data/csse_covid_19_time_series/"
file_names = c("time_series_covid19_confirmed_global.csv",
"time_series_covid19_deaths_global.csv")

urls= str_c(url_in, file_names)
global_cases = read_csv(urls[1])
global_death = read_csv(urls[2])


# Convert global cases and death dataset to a longer format while removing Lat and Long columns
global_cases = global_cases %>% pivot_longer(cols = -c('Province/State','Country/Region',Lat, Long), names_to = "date", values_to = "cases")%>% select(-c(Lat, Long))

global_death = global_death %>% pivot_longer(cols = -c('Province/State','Country/Region', Lat, Long), names_to = "date", values_to = "deaths")%>% select(-c(Lat, Long))

#Summary of both dataset
summary(global_cases)
summary(global_death)

#Combine both dataset using a full join 
#Renaming the County/Region and Province State column to "Country_Region" & "Province_State" respectively
#Change the date column to a date Object using Ludridate function mdy()
global = global_cases %>% full_join(global_death) %>% rename(Country_Region = 'Country/Region', Province_State = 'Province/State')%>% mutate(date=mdy(date))

#Summary, tail & head of data
summary(global)
head(global, 20)
tail(global)

#filtering data by cases above zero.
global = global %>% filter(cases > 0) 
summary(global)

#combine Province_State & Country_Region into a single column "COmbined_key" seperated by a comma
global = global %>% unite("Combined_key", c(Province_State, Country_Region), sep = ",", na.rm = TRUE, remove = FALSE)

#read in population data off the John Hopkin website while removing columns not needed
uid_lookup_url ="https://raw.githubusercontent.com/CSSEGISandData/COVID-19/master/csse_covid_19_data/UID_ISO_FIPS_LookUp_Table.csv"
uid = read_csv(uid_lookup_url) %>% select(-c(Lat, Long_, Combined_Key, code3, iso2, iso3, Admin2))
uid

#Conduct a left join of the global dataset with the population data
#Removing Rpovince_state and Combined_key
global_join = global %>% left_join(uid, by = c("Province_State", "Country_Region")) %>%
select(-c(UID, FIPS)) %>% select (Country_Region, date, cases, deaths, Population)

global_join

#fragment the date object to get the year and month and saved to new columns
global_frag = global_join %>% mutate(year = year(date), month = month(date))

#group covid cases by countries and their cases/deaths per million population
#drop rows with NA
global_issue = global_join %>% select(-c(date)) %>% group_by(Country_Region)%>% summarize(cases=sum(cases),deaths=sum(deaths), Population = mean(Population))%>% mutate(deaths_per_mill = deaths*1000000/Population, cases_per_mill = cases*1000000/Population
)%>% select(Country_Region, cases, deaths, deaths_per_mill, cases_per_mill, Population) %>% ungroup() %>% drop_na()

#group covid cases by countries and their cases/deaths per thousand population
#filter columns to show rows with population and cases greater than 0
global_issue2 = global_issue %>% group_by(Country_Region)%>%summarize(deaths=max(deaths), cases = max(cases), population = max(Population),cases_per_thou=1000*cases/population, deaths_per_thou = 1000*deaths/population) %>% filter(cases > 0, population >0)

##10 countries with least deaths per thoudand from Covid-19
global_issue2 %>% slice_min(deaths_per_thou, n = 10)

##0 10 countries with greater deaths per thoudand from Covid-19
global_issue2 %>% slice_max(deaths_per_thou, n = 10)

#Checking distribution of covid19 cases and deaths in Nigeria per from 2020 till when the data ends
Country= "Nigeria"
global_frag %>% filter(Country_Region == Country) %>% ggplot(aes(x=date, y = cases)) + geom_line(aes(color = "cases")) + geom_point(aes(color = "cases")) + geom_line(aes(y = deaths, color = "deaths")) + geom_point(aes(y = deaths, color = "deaths")) + scale_y_log10()+ theme(legend.position="bottom", axis.text.x= element_text(angle = 90)) + labs(title = str_c("COVID19 in ", Country), y=NULL)

#Summarize Nigeria data by cases/deaths per million
Nigeria_summary = global_frag %>% filter(Country_Region == Country) %>% select(-c(month, year)) %>% mutate(cases_per_mill = cases*1000000/Population,deaths_per_mill = deaths*1000000/Population)%>% select(Country_Region, date, cases, deaths, cases_per_mill, deaths_per_mill, Population)
Nigeria_summary

#Summarize Nigeria data by yearand cases/deaths per million
Nigeria_year_summary = global_frag %>% filter(Country_Region == Country) %>% select(-c(month)) %>%
group_by(Country_Region,year) %>% summarize(cases=sum(cases),deaths=sum(deaths), Population = mean(Population))%>% mutate(cases_per_mill = cases*1000000/Population,deaths_per_mill = deaths*1000000/Population)%>% select(Country_Region, year, cases, deaths, cases_per_mill, deaths_per_mill, Population) %>% ungroup()
Nigeria_year_summary

#plotting totals deaths/cases per million each year in Nigeria
Nigeria_year_summary%>% ggplot(aes(x=year, y = cases)) + geom_line(aes(color = "cases")) + geom_point(aes(color = "cases")) + geom_line(aes(y = deaths, color = "deaths")) + geom_point(aes(y = deaths, color = "deaths")) + scale_y_log10()+ theme(legend.position="bottom", axis.text.x= element_text(angle = 90)) + labs(title = str_c("COVID19 in ", Country), y=NULL)

#Summarize Nigeria data by month and cases/deaths per million
Nigeria_month_summary = global_frag %>% filter(Country_Region == Country) %>% select(-c(year)) %>%
group_by(Country_Region,month) %>% summarize(cases=sum(cases),deaths=sum(deaths), Population = mean(Population))%>% mutate(cases_per_mill = cases*1000000/Population,deaths_per_mill = deaths*1000000/Population)%>% select(Country_Region, month, cases, deaths, cases_per_mill, deaths_per_mill, Population) %>% ungroup()
Nigeria_month_summary

#plotting totals deaths/cases per million each month in Nigeria
Nigeria_month_summary%>% ggplot(aes(x=month, y = cases)) + geom_line(aes(color = "cases")) + geom_point(aes(color = "cases")) + geom_line(aes(y = deaths, color = "deaths")) + geom_point(aes(y = deaths, color = "deaths")) + scale_y_log10()+ theme(legend.position="bottom", axis.text.x= element_text(angle = 90)) + labs(title = str_c("COVID19 in ", Country), y=NULL) + scale_x_continuous(breaks = seq(1, 12, 1))

#filter global issues to know total deaths per million and cases per million in Nigeria
Nigeria_issue = global_issue %>% filter(Country_Region == Country)
Nigeria_issue

#Modeling your global cases per million to deaths per million using a linear model
mod1 = lm(deaths_per_mill ~ cases_per_mill , data = global_issue)
summary(mod1)

#Extract R-squared and coefficients
r_squared = summary(mod1)$r.squared
coefficients = summary(mod1)$coefficients
intercept = coefficients[1, 1]
slope = coefficients[2, 1]

#Add new column into global_issue and save to new variable "global_pred"

global_pred = global_issue %>% mutate(pred=predict(mod1))

#Plot linear model along with R2 and equation
global_pred %>% ggplot() +
geom_point(aes(x=cases_per_mill, y=deaths_per_mill),color="blue") + 
geom_line(aes(x=cases_per_mill, y=pred),color="red") +labs(title = "Linear model of global cases per million to deaths per million") + annotate("text", x = Inf, y = Inf, label = paste("R-squared:", round(r_squared, 2), "\nEquation: y =", round(intercept, 1), "+", round(slope, 1), "* x"), hjust = 1.1, vjust = 2, size = 4, color = "black")

#Modeling your Nigeria cases per million to deaths per million using a linear model
mod = lm(deaths_per_mill ~ cases_per_mill , data = Nigeria_summary)
summary(mod)

#Extract R-squared and coefficients
r_squared1 =  summary(mod)$r.squared
coefficients1 = summary(mod)$coefficients
intercept1 = coefficients[1, 1]
slope1 = coefficients[2, 1]

#Add new column into Nigeria_issue and save to new variable "Nigeria_pred"
Nigeria_pred = Nigeria_summary %>% mutate(pred=predict(mod))

#Plot linear model along with R2 and equation
Nigeria_pred %>% ggplot() +
geom_point(aes(x=cases_per_mill, y=deaths_per_mill),color="blue") + 
geom_line(aes(x=cases_per_mill, y=pred),color="red") +labs(title = "Linear model of Nigeria cases per million to deaths per million") + annotate("text", x = Inf, y = Inf, label = paste("R-squared:", round(r_squared1, 2), "\nEquation: y =", round(intercept1, 1), "+", round(slope1, 1), "* x"), hjust = 2, vjust = 2, size = 4, color = "black")

```

**Conclusion:** Given all my analysis and modelling, I have come to the following conclusions below;-

*Nigeria Covid-19 cases and death have slightly increased cases and deaths in January to March likely due to increased festive activity during that period leading to  more contact with population*. *There is a nearly perfect relationship between the no of covid cases per million and deaths per million.*

**Possible sources of Bias:** I am personally biased toward focusing the analysis on Nigeria as I come from Nigeria. In addition, the prediction of Covid 19 deaths in Nigeria could be more precise if other information about the cases such as age, sex, gender, medical history e.t.c. was available. 