---
output:
  github_document: default
  html_document: default
---
<<<<<<< HEAD
---
title: "NYPD"
author: "G. Ubit"
date: "2025-05-07"
output: github_document
---
```{r, echo = FALSE}
knitr::opts_chunk$set(
  fig.path = "README_figs/README-")
```
Statement of Interest: I am interested in understanding  the number of crime rate across different borough in New York City.In addition, I am curious about the timing of these crimes particularly in Bronx.

Dataset summary: The dataset represent a breakdown of shooting incidents in the 5 boroughs of NY city between 2006 to present. Details per incident include event, location, time of occurrence, details of victims and perpetrators etc. More details can be found at <https://catalog.data.gov/dataset/nypd-shooting-incident-data-historic>


```{r}
#library needed for analysis
library(lubridate)
library(tidyverse)
library(tidymodels)
library(dplyr)
library(ggplot2)
library(readr)
library(utils)
library(patchwork)

##importing data from the city of New York website and saving to a variable
url = "https://data.cityofnewyork.us/api/views/833y-fsy8/rows.csv?accessType=DOWNLOAD"

NYPD_main = read_csv(url)

# Check the summary of the Data
summary(NYPD_main)

```{r}
# Retain needed columns while changing OCCUR date to a new coumn name/date object
NYPD_select = NYPD_main %>% mutate(date=mdy(OCCUR_DATE)) %>% select(c(date, OCCUR_TIME, BORO, LOC_OF_OCCUR_DESC,PERP_AGE_GROUP,PERP_AGE_GROUP, PERP_SEX,  PERP_RACE, VIC_AGE_GROUP,VIC_SEX, VIC_RACE, INCIDENT_KEY))

# OOps, seems we have some NAs in some columns - Removing more columns and renaming some of the remaining column
NYPD_select = NYPD_select %>% rename(District = BORO, Victim_Age_Group = VIC_AGE_GROUP, Victim_Sex = VIC_SEX, Victim_race=VIC_RACE, 
time = OCCUR_TIME) %>% select(-c(LOC_OF_OCCUR_DESC,PERP_AGE_GROUP,PERP_SEX,PERP_RACE))

#Sort data by date and time and row duplicate rows
NYPD_select = NYPD_select%>% arrange(date, time) %>%distinct()
NYPD_select
#How many district are avaliable
n_distinct(NYPD_select$District)

#disctint count of victim's race per district
NYPD_select%>% group_by(District)%>%summarize(distinct_race=n_distinct(Victim_race))


#No of crime per district and sorted using distinct crime
crime = NYPD_select%>% group_by(District)%>%summarize(distinct_crime=n_distinct(INCIDENT_KEY))%>% arrange(distinct_crime)

#No of crime per district and time
Crime_Date = NYPD_select%>% group_by(District, time)%>%summarize(distinct_crime=n_distinct(INCIDENT_KEY))
Crime_Date

#plot graph of total no of crime per district
crime %>% ggplot(aes(x = District, y = distinct_crime, fill=District)) +
  geom_col() +labs(title= "Crime per Districts of New York City")


#Declare region of interest
district1 = "BRONX"
district2 = "MANHATTAN"

# Filter and plot crimes per hours in Bronx and Manhattan individually
Crime_Date %>%
  filter(District == district1) %>%
  ggplot(aes(x = time, y = distinct_crime, fill = time)) +
  geom_line() +
  coord_cartesian(ylim = c(0, max(Crime_Date$distinct_crime, na.rm = TRUE))) +
  labs(title = paste("Crime per minute in", district1), x = "Time in h:m:s", y = "Distinct Crime") +
  theme_minimal() +

Crime_Date %>%
  filter(District == district2) %>%
  ggplot(aes(x = time, y = distinct_crime, fill = time)) +
  geom_line() +
  coord_cartesian(ylim = c(0, max(Crime_Date$distinct_crime, na.rm = TRUE))) +
  labs(title = paste("Crime per minute in", district2), x = "Time in h:m:s", y = "Distinct Crime") +
  theme_minimal() +
  scale_fill_viridis_c()

#filter data for Bronx and Manhattan and save to a variable
Bronx_crime= Crime_Date %>% filter(District == district1)
Manhattan_crime= Crime_Date %>% filter(District == district2)

# Quadratic model for district1
model_district1_quad <- lm(distinct_crime ~ poly(time, 5), Bronx_crime)
# Quadratic model for district2
model_district2_quad <- lm(distinct_crime ~ poly(time, 2), Manhattan_crime)

# Summary of the model for Bronx_crime
summary(model_district1_quad)

# Summary of the model for Manhattan_crime
summary(model_district2_quad)

#Add model as a column to Bronx_crime and Manhattan_crime data
Bronx_crime = Bronx_crime %>% mutate(pred = predict(model_district1_quad))
Manhattan_crime = Manhattan_crime %>% mutate(pred = predict(model_district2_quad))

# Plot for Bronx quadratic regression line
Crime_Date %>%
  filter(District == district1) %>%
  ggplot(aes(x = time, y = distinct_crime)) +
  geom_line() +
  geom_smooth(method = "lm", formula = y ~ poly(x, 2), se = FALSE, color = "blue") +
  coord_cartesian(ylim = c(0, max(Crime_Date$distinct_crime, na.rm = TRUE))) +
  labs(title = paste("Crime per minute in", district1), x = "Time in h:m:s", y = "Distinct Crime") +
  theme_minimal()

# Plot for Manhattan with quadratic regression line

Crime_Date %>%
  filter(District == district2) %>%
  ggplot(aes(x = time, y = distinct_crime)) +
  geom_line() +
  geom_smooth(method = "lm", formula = y ~ poly(x, 2), se = FALSE, color = "blue") +
  labs(title = paste("Crime per minute in", district2), x = "Time in h:m:s", y = "Distinct Crime") +
  theme_minimal()
```

#Conclusion
The data showed that crime were mostly carried in late at night and in the early hours of the morning. In addition, Brookyln has the highest amount of crime during the period of time under investigation followed by Bronx, Queens, Manhattan and Staten Island.
# Sources of Bias
1. The model is biased towards outliers.
2. There is a lot of missing information about the perpetrators, causing inadequate analysis on their sex, race, and age group.
=======
---
title: "NYPD"
author: "G. Ubit"
date: "2025-05-07"
output: github_document
---
```{r, echo = FALSE}
knitr::opts_chunk$set(
  fig.path = "README_figs/README-"
)
```
Statement of Interest: I am interested in understanding  the number of crime rate across different borough in New York City.In addition, I am curious about the timing of these crimes particularly in Bronx.

Dataset summary: The dataset represent a breakdown of shooting incidents in the 5 boroughs of NY city between 2006 to present. Details per incident include event, location, time of occurrence, details of victims and perpetrators etc. More details can be found at <https://catalog.data.gov/dataset/nypd-shooting-incident-data-historic>

```{r}
#library needed for analysis
library(lubridate)
library(tidyverse)
library(tidymodels)
library(dplyr)
library(ggplot2)
library(readr)
library(utils)
library(patchwork)

##importing data from the city of New York website and saving to a variable
url = "https://data.cityofnewyork.us/api/views/833y-fsy8/rows.csv?accessType=DOWNLOAD"

NYPD_main = read_csv(url)

# Check the summary of the Data
summary(NYPD_main)

```{r}
# Retain needed columns while changing OCCUR date to a new coumn name/date object
NYPD_select = NYPD_main %>% mutate(date=mdy(OCCUR_DATE)) %>% select(c(date, OCCUR_TIME, BORO, LOC_OF_OCCUR_DESC,PERP_AGE_GROUP,PERP_AGE_GROUP, PERP_SEX,  PERP_RACE, VIC_AGE_GROUP,VIC_SEX, VIC_RACE, INCIDENT_KEY))

# OOps, seems we have some NAs in some columns - Removing more columns and renaming some of the remaining column
NYPD_select = NYPD_select %>% rename(District = BORO, Victim_Age_Group = VIC_AGE_GROUP, Victim_Sex = VIC_SEX, Victim_race=VIC_RACE, 
time = OCCUR_TIME) %>% select(-c(LOC_OF_OCCUR_DESC,PERP_AGE_GROUP,PERP_SEX,PERP_RACE))

#Sort data by date and time and row duplicate rows
NYPD_select = NYPD_select%>% arrange(date, time) %>%distinct()
NYPD_select
#How many district are avaliable
n_distinct(NYPD_select$District)

#disctint count of victim's race per district
NYPD_select%>% group_by(District)%>%summarize(distinct_race=n_distinct(Victim_race))


#No of crime per district and sorted using distinct crime
crime = NYPD_select%>% group_by(District)%>%summarize(distinct_crime=n_distinct(INCIDENT_KEY))%>% arrange(distinct_crime)

#No of crime per district and time
Crime_Date = NYPD_select%>% group_by(District, time)%>%summarize(distinct_crime=n_distinct(INCIDENT_KEY))
Crime_Date

#plot graph of total no of crime per district
crime %>% ggplot(aes(x = District, y = distinct_crime, fill=District)) +
  geom_col() +labs(title= "Crime per Districts of New York City")


#Declare region of interest
district1 = "BRONX"
district2 = "MANHATTAN"

# Filter and plot crimes per hours in Bronx and Manhattan individually
Crime_Date %>%
  filter(District == district1) %>%
  ggplot(aes(x = time, y = distinct_crime, fill = time)) +
  geom_line() +
  coord_cartesian(ylim = c(0, max(Crime_Date$distinct_crime, na.rm = TRUE))) +
  labs(title = paste("Crime per minute in", district1), x = "Time in h:m:s", y = "Distinct Crime") +
  theme_minimal() +

Crime_Date %>%
  filter(District == district2) %>%
  ggplot(aes(x = time, y = distinct_crime, fill = time)) +
  geom_line() +
  coord_cartesian(ylim = c(0, max(Crime_Date$distinct_crime, na.rm = TRUE))) +
  labs(title = paste("Crime per minute in", district2), x = "Time in h:m:s", y = "Distinct Crime") +
  theme_minimal() +
  scale_fill_viridis_c()

#filter data for Bronx and Manhattan and save to a variable
Bronx_crime= Crime_Date %>% filter(District == district1)
Manhattan_crime= Crime_Date %>% filter(District == district2)

# Quadratic model for district1
model_district1_quad <- lm(distinct_crime ~ poly(time, 5), Bronx_crime)
# Quadratic model for district2
model_district2_quad <- lm(distinct_crime ~ poly(time, 2), Manhattan_crime)

# Summary of the model for Bronx_crime
summary(model_district1_quad)

# Summary of the model for Manhattan_crime
summary(model_district2_quad)

#Add model as a column to Bronx_crime and Manhattan_crime data
Bronx_crime = Bronx_crime %>% mutate(pred = predict(model_district1_quad))
Manhattan_crime = Manhattan_crime %>% mutate(pred = predict(model_district2_quad))

# Plot for Bronx quadratic regression line
Crime_Date %>%
  filter(District == district1) %>%
  ggplot(aes(x = time, y = distinct_crime)) +
  geom_line() +
  geom_smooth(method = "lm", formula = y ~ poly(x, 2), se = FALSE, color = "blue") +
  coord_cartesian(ylim = c(0, max(Crime_Date$distinct_crime, na.rm = TRUE))) +
  labs(title = paste("Crime per minute in", district1), x = "Time in h:m:s", y = "Distinct Crime") +
  theme_minimal()

# Plot for Manhattan with quadratic regression line

Crime_Date %>%
  filter(District == district2) %>%
  ggplot(aes(x = time, y = distinct_crime)) +
  geom_line() +
  geom_smooth(method = "lm", formula = y ~ poly(x, 2), se = FALSE, color = "blue") +
  labs(title = paste("Crime per minute in", district2), x = "Time in h:m:s", y = "Distinct Crime") +
  theme_minimal()
```

#Conclusion
The data showed that crime were mostly carried in late at night and in the early hours of the morning. In addition, Brookyln has the highest amount of crime during the period of time under investigation followed by Bronx, Queens, Manhattan and Staten Island.
# Sources of Bias
1. The model is biased towards outliers.
2. There is a lot of missing information about the perpetrators, causing inadequate analysis on their sex, race, and age group.
>>>>>>> 096b965cbdc855b955c7fc66ad6baeacfc6b532a
